---
title: 'Lab #4'
author: "Nikole Slinger"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1.

Although the three Excel sheets are loaded into R, they are not useful before some cleaning. Manipulate the three imported data frames, so that in the cleaned dataset:

1. The player names and opponent names are factors.
2. All offensive and defensive statistics are numerical. The Weight column in biogrpahy is also numeric. (Optional: Read ?across and learn to mutate multiple columns simultaneously using the across function in dplyr)
3. Change the Height column in biography into numeric. (Hint: Use a unit that works well)

```{r}

```

## Part 2.

### Problem 1.
Reformat the offClean data frame into a tidy format using pivot_longer. The type of statistic (Receiving_REC for the number of receivings, Receiving_YDS for receiving yards) is added as a new key column named stat.

```{r}

```

### Problem 2.
Show on the distributions of the offensive statistics. The distribution of each offensive statsitic should be shown in a separate facet. Explain the patterns you see.

```{r}

```

### Problem 3.
Did ISU have better offense against Oregon or Oklahoma? Answer this question by creating a scatterplot with the x- and y-axes being the receiving yards of the ISU players against these two opponents. A large number of receiving yards indicates good offense.

```{r}

```

### Problem 4.
Separate the Hometown column into two columns, namely, the city and the state where the player comes from. Print the first few rows of these two columns. (Hint: look at the sep= argument of separate)

```{r}

```

### Problem 5.
How many players are there from each state?

```{r}

```

## Part 3.

### Problem 1.
Find which states do the ISU players (both defensive and offensive) come from. Create an appropriate numerical or graphical summary.

```{r}

```

### Problem 2.
How does the performance of quarterback Brock Purdy relate to the performance of the ISU team?

```{r}

```

### Problem 3.
Which ISU player(s) made the biggest progress over the last year? Compare the player-level statistics (i.e. averages and totals for each player) using both the 2020 and the 2019 data here. Choose for yourself the player statistics to investigate.

```{r}

```


