---
title: "Lab #4 Team"
author: "Julia Lundstrum, Nikole Slinger, and Zack Larson"
output: html_document
---

# Loading Data
```{r}
library(tidyverse)
library(readxl)
defense <- read_excel('./cyclonesFootball2020.xlsx', sheet='Defensive')
offense <- read_excel('./cyclonesFootball2020.xlsx', sheet='Offensive')
biography <- read_excel('./cyclonesFootball2020.xlsx', sheet='Biography')
```

# Part One
Here we've successfully made the player and opponent name columns into factors. We then make all statistics numerical and separate columns as needed.
```{r}
# offense
offClean <- offense %>%
  mutate(across(c('Name','Opponent_Opponent'), as.factor)) %>%
  separate(`Passing_CMP-ATT`, c('Passing_CMP','Passing_ATT')) %>%
  mutate(across(c(3:13), as.numeric))
str(offense)
# defense
defClean <- defense %>%
  mutate(across(c('Name','Opponent_Opponent'), as.factor)) %>%
  mutate(across(c(3:11), as.numeric))
str(defense)
# biography
bioClean <- biography %>%
  mutate(Name = as.factor(Name)) %>%
  separate(Height, c('Ft', 'In')) %>%
  mutate(across(c('Ft','In', 'Weight'), as.numeric), Height = 12 * Ft + In)
#Separated height and made new height column that is the total height in inches.
str(biography)
```

# Part Two
### 1.
Using `pivot_longer` to reformat the data set and include a new column.
```{r}
offCleanp <- pivot_longer(offClean, Receiving_REC:Passing_INT, names_to='stat', values_to='Amount') 
```

### 2.
This data shows that there are some stats that have a wide distribution among players while others are much more smaller in distribution. This can mainly be because of the typical number of players who usually obtain these stats or based on the "number"/value of the stat itself. For example, the passing touchdowns stat has a large range of stat accumulation because there is typically only one quarterback who throws the passes on a team in a game. On the other hand, receiving yards is more widely distributed because there are several relievers who catch a ball and accumulate receiving yards in a game. Additionally, Passing, Receiving, and Rushing Yards (the three most distributed stats) have values such as 200 or higher, because it makes sense for players to have a lot of "yards", where on the other hand the intercept stats are much lower because intercepts don't happen too often in games, usually only once or twice.

```{r, fig.width=10, fig.height=6}
# plotting a facet wrap with the stats
ggplot(subset(offCleanp, !is.na(Amount)), aes(y = Amount, colour = factor(stat))) + geom_point(stat='count') + facet_wrap(. ~ stat, scales="free") + xlab("Number of Players") + theme(axis.text.x = element_text(angle = 90))
```

### 3.
When based off of the Receiving Yards stat, ISU Football looked to have a better game offensively against Oklahoma because there were more points towards the right side of the scatterplot meaning there were many more receiving yards. There is also a higher distribution of more receiving yards in general against Oklahoma compared to Oregon. The boxplot confirms what the scatterplot suggests, showing a distribution of more receiving yards than in the game against Oklahoma. Also, although the median value seems to be the same for both opponents, the range extends much higher in the Oklahoma game.

```{r}
# separating so we only have Oregon and Oklahoma Receiving Yard data
goodOffense <- offCleanp %>% 
  filter(stat == "Receiving_YDS" & (Opponent_Opponent == "Oklahoma" | Opponent_Opponent == "Oregon") & !is.na(Amount))
# creating scatter plot and boxplot
ggplot(goodOffense, aes(y = Amount, colour = Opponent_Opponent, shape = Opponent_Opponent)) + geom_point(stat='count') + xlab("Number of Players") + ylab("Receiving Yards")

boxplot(data = goodOffense, Amount~Opponent_Opponent, main = 'Best ISU Offensive Game: Oklahoma or Oregon', 
        xlab = 'Team', ylab = 'Receiving Yards')
```

### 4.
We separated `Hometown` into `City` and `State` using the ', ' characters.
```{r}
biography <- separate(bioClean, `Hometown`, c('City', 'State'), sep=', ')
head(biography)
```

### 5.
Based on the output, we can can read a few of the rows. For example, there are 42 players from Iowa, 4 players from Arizona, and even 1 player from Australia.
```{r}
biography %>% count(State)
```

# Part Three
### 1.
Based off of both numerical and graphical summaries, we can tell that the most frequent home state of offensive players at Iowa State is Florida with 3 players, while the most frequent home state for defensive players is Iowa with 10 players.

```{r}
# using inner join to get only offensive player names and states
OPlayers <- inner_join(biography, offClean, by = 'Name')
OPlayers <- OPlayers %>% select(Name, State)
OPlayers <- OPlayers %>% group_by(Name) %>% distinct()
OPlayers
# using inner join to get only defensive player names and states
DPlayers <- inner_join(biography, defClean, by = 'Name')
DPlayers <- DPlayers %>% select(Name, State)
DPlayers <- DPlayers %>% group_by(Name) %>% distinct()
DPlayers

# making states into factors for bar chart
DPlayers$State <- factor(DPlayers$State)
OPlayers$State <- factor(OPlayers$State)

# creating bar chart for offensive and defensive data
ggplot(OPlayers, aes(x = State)) + geom_bar(fill = 'blue') + 
  ggtitle('ISU Offense Player Home States') + xlab('State') + ylab('Count')
ggplot(DPlayers, aes(x = State)) + geom_bar(fill = 'blue') + 
  ggtitle('ISU Defense Player Home States') + xlab('State') + ylab('Count')
``` 

### 2.
For our analysis, we looked at the total yards that Purdy accumulated via rushing and passing, then compared those stats with the outcome of the game (either a win or loss). Zack found the win and loss data easily online and added that to the existing Purdy data frame. We then found that Purdy's performance does seem to relate to the Iowa State's performance as a team because in the graph shown, Purdy has the most total yards in the games where Iowa State wins. Looking closer, we can see that where there are clearly more than 200 total yards, Iowa State almost always won, while in 2 of the games where ISU lost (Louisiana and Oklahoma State), Purdy had noticeably fewer total yards. The exception to this is the second game against Oklahoma where Purdy had the highest number of total yards (300+), but ISU still lost.

```{r}
# creating total yards column to use as purdy's primary stat
Purdy <- offClean %>% filter(Name == "Purdy, Brock")
Purdy <- Purdy %>% mutate(TotalYards = Passing_YDS + Rushing_YDS)
Purdy <- Purdy %>% select(Name, Opponent_Opponent, TotalYards)

#Found game results for each opponent online and added them to df to see team performance vs Purdy's performance
Purdy$GameResult <- c("Loss", "Win", "Win", "Win", "Loss", "Win", "Win", "Win", "Win", "Win", "Loss", "Win")
ggplot(Purdy, aes(x = factor(Opponent_Opponent), y = TotalYards,fill=GameResult)) +
  geom_histogram(stat = 'Identity', position = 'dodge') + ggtitle("Brock Purdy Statistics") +
  ylab("Total Yards") + xlab("Opponent") + theme(axis.text.x = element_text(angle = 90))
```

### 3.
We decided to focus the study of improvement on defensive players, more specifically looking at the average number of solo tackles in each game. We downloaded the 2019 data and compared that data to the current season's data (2020) and found the 15 players who played in both years. Then we ordered them by the average number of tackles that they gained (or lost) per game from last year. In doing so, we found that Mike Rose, Jake Hummel, and JaQuan Bailey were among the most improved players on defense because they all got, on average, over one more solo tackle a game than the previous year (specifically 1.5, 1.41, and 1.17 more tackles, respectively).

```{r}
#Compare defense players' tackles after cleaning the 2019 data
defense2019 <- read_excel('./cyclonesFootball2019.xlsx', sheet = 'Defensive')

# fixing column names so the names are factors and stats are numeric
defense2019 <- defense2019 %>%
  mutate(across(c('Name','Opponent_Opponent'), as.factor),
         across(Tackles_Solo:Pass_PB, as.numeric))
# 2020 summaries
defense2020 <- defClean %>% 
  group_by(Name) %>% distinct() %>% 
  summarise(soloTacklesFrom2020 = mean(Tackles_Solo, na.rm = TRUE))
# 2019 summaries
defense2019 <- defense2019 %>% 
  group_by(Name) %>% distinct() %>%
  summarise(soloTacklesFrom2019 = mean(Tackles_Solo, na.rm = TRUE))
# joining summaries into new data frame to compare
defenseImprovements <- inner_join(defense2020, defense2019, by = "Name")
# finding differences between 2019 and 2020
defenseImprovements <- defenseImprovements %>%
  mutate(numTacklesIncreased = soloTacklesFrom2020 - soloTacklesFrom2019)
# sorting by most improved
defenseImprovements[order(defenseImprovements$numTacklesIncreased, decreasing = TRUE),]
```
